---
title: "SST Analysis"
author: "Aaron Shavitz"
format: 
  html:
    code-fold: true
    code-tools: true
    code-summary: "Show code"
---

```{r, echo=FALSE}
# Using knitr package to read code chunks from sst_anaysis.R

# Set the root directory to the current folder
knitr::opts_knit$set(root.dir = here::here())

# Read in the R script from the 'scripts' folder (relative to the root project directory)
knitr::read_chunk("../scripts/sst_analysis.R")
```

### SST Impact on Northeastern US Eelgrass Cover - Analysis
Data source details:
```{r Header}
```

### Load Libraries
```{r Load Libraries, message=FALSE, warning=FALSE}
```

### Set Global ggplot Themes
```{r Set Global ggplot Themes, message=FALSE, warning=FALSE}
```

### Load and Organize SST Data
```{r Load and Organize SST Data, message=FALSE, warning=FALSE}
```

### Load and Organize Data from Plaisted et al. 2022
```{r Load and Organize Data from Plaisted et al. 2022, message=FALSE, warning=FALSE}
```

### Compare CoastWatch SST to Plaisted et al. 2022 Temperature Data
```{r Compare CoastWatch sst to Plaisted et al. 2022 Temperature Data, message=FALSE, warning=FALSE}
```


THINGS ARE MISSING

### Prepare Data for Causal Modeling
```{r Prepare Data for Causal Modeling, message=FALSE, warning=FALSE}
```

### Aggregated Percent Cover Model
```{r Aggregated Percent Cover Model, message=FALSE, warning=FALSE}
```

### Quadrat Presence Absence Model
```{r Quadrat Presence Absence Model, message=FALSE, warning=FALSE}
```

### Quadrat Percent Cover *Given Coverage* Model
```{r Quadrat Percent Cover *Given Coverage* Model, message=FALSE, warning=FALSE}
```