---
title: "My Example Talk"
# bibliography: test_library.bib
title-slide-attributes: 
  data-background-color: "blue"
  data-background-opacity: "50%"
author: "Jarrett Byrnes"
format: 
  revealjs:
      theme: moon
      transition: slide
---

## Here is my example talk

This is a slide

# This is a section break slide


## This is another slide

Hello!

## An Incremental List Slide

:::{.incremental}
- This is the first thing
- This is the second thing
:::

## A slide with a pause
```{r}
library(palmerpenguins)
data(penguins)
boxplot(body_mass_g ~ species, data = penguins)
```

. . .

Those are some big penguins!

## Columns!

::::: {.columns}

:::: {.column width="40%"}
::: {.incremental}
- this is my first point about this figure. 
- This is my second point about this figure
:::
::::


:::: {.column width="60%"}
```{r}
library(palmerpenguins)
data(penguins)
boxplot(body_mass_g ~ species, data = penguins)
```
::::


:::::

## Sometimes you have a lot of text it might be hard to fit on a slide {.smaller}

- right
- right
- right
- right
- right
- right
- right


## Speaker Notes

This is a slide

::: {.notes}
These are my speaker notes
:::

## Here is a slide with a footer
```{r}
library(palmerpenguins)
data(penguins)
boxplot(body_mass_g ~ species, data = penguins)
```

::: footer
Data supplied by the Palmer Station LTER
:::


## A Different Slide background color {background-color="blue"}

This is a blue slide

## {background-image="https://i.imgflip.com/2tuc2c.jpg" background-size="contain"}

## This is a referenced slide

This is a slide [@li_effects_2015]

## References

::: {#refs}
:::
